\chapter{Conclusion}
\label{ch:Conclusion}
In this work we introduced a novel approach to solving quantifier free bitvector problems in SMT-LIB's \texttt{QF\_BV} theory. The approach is based on abstraction methodologies previously used for various other problems in logic and specifically in SMT.
On the one hand we presented numerous abstractions for 3 comparatively costly funcitons of the bitvector theory, on the other hand we proposed a simple theoretical framework allowing a proof of correctness for the presented abstractions.
While we saw that the presented approach performs better than current state-of-the-art solvers in deciding unsatisfiable bitvector problems, the implemented prototype is not yet competitive for satisfiable instances. This is of course in some way a natural result, as over-approximations usually improve the solver runtime on unsatisfiable (and not on satisfiable) instances.

\paragraph{Future Work}
With the abstraction's correctness proved and the abstraction's performance evaluated through the current prototype, one could now implement the abstraction scheme directly into a SMT solver like Boolector making use of interleaved under- and over-approximations. We believe that this might enhance the solver's performance sufficiently to be competitive for both satisfiable and unsatisfiable instances.
Additionally, a time limit for each refinement round could be introduced in order to avoid cases where the solver gets stuck in some specific refinement step. For this, a detailed parameter analysis will be necessary to find a time limit that keeps the abstraction steps effective while avoiding dead ends. Finally, it is left to investigate whether a don't care reasoning strategy similar to the one used for lemmas on demand in Boolector \cite{NiemetzPreinerBiere-FMCAD14} could improve the abstraction refinement procedure's performance.\\
In any case, using abstractions in solving bitvector SMT problems seems like a promising pathway that most certainly will be further investigated in future work.